---
title: "EDA and Modeling of Health Data"
author: "Ada Lazuli"
date: '2022-06-20'
output:
  html_document:
    toc: true
    toc_depth: 2
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("~/Desktop/DATA630/assignment2/")
library(ggplot2)

outliers <- function(x) {
  # Find the outliers  by filtering values less that 3 STD from the mean or 3 STD greater than the mean
  min <- mean(x)  - 3 * sd(x) 
  max <- mean(x) + 3 * sd(x)
  
  return(x[x < min | x > max])
}

```

# Data Exploration Of [Heart Data](https://learn.umgc.edu/content/enforced/685926-027339-01-2225-GO1-9041/Dataset%20Descriptions/whas500.txt)

## Initial Data Inspection

```{r}
df <- read.csv("whas500.csv")
str(df)
```

The data consists of `r nrow(df)` observations, `r ncol(df)` variables, and exactly `r sum(is.na(df))` missing values.

The id field can be dropped from the data set, as it does not provide useful information.

```{r}
df <- df[, -1]
```

Next, the admittance date, discharge date, and the follow up date can be removed 

```{r}
df <- df[, c(-15,-16,-17)]
```

Finally remove the length of hospital stay, total length of follow-up, and vital status at the last follow-up

```{r}
df <- df[, c(-15, -17, -18)]
colnames(df)
```

### Age

The age is a discrete quantitative variable that describes the age of the individual when they were admitted to the hospital.

```{r}
summary(df$age)
ggplot(df, aes(x = age)) + geom_histogram() + geom_vline(aes(xintercept = mean(df$age)), color = "red")
```

There were `r length(outliers(df$age))` outliers found in the data for age.

### gender

The gender variable refers to the gender of the individual admitted to the hospital.
```{r}
print(table(df$gender))
```

###  hr (Initial Heart Rate)

The initial heart rate of the individual admitted to the hospital

```{r}

summary(df$hr)
ggplot(df, aes(x = hr)) + geom_histogram() + geom_vline(aes(xintercept = mean(df$hr)), color = "red")
```

There were `r length(outliers(df$hr))` outliers found in the data for hr. The outliers are `r outliers(df$hr)` for the initial heart rate.


### sysbp (Systolic Blood Pressure)

The Systolic Blood Pressure of the individual admitted to the hospital

```{r}

summary(df$sysbp)
ggplot(df, aes(x = sysbp)) + geom_histogram() + geom_vline(aes(xintercept = mean(df$sysbp)), color = "red")
```

There were `r length(outliers(df$sysbp))` outliers found in the data for hr. The outliers are `r outliers(df$sysbp)` for the Systolic Blood Pressure.


### diasbp (Diastolic Blood Pressure)

The Diastolic Blood Pressure of the individual admitted to the hospital

```{r}

summary(df$diasbp)
ggplot(df, aes(x = diasbp)) + geom_histogram() + geom_vline(aes(xintercept = mean(df$diasbp)), color = "red")
```

There were `r length(outliers(df$diasbp))` outliers found in the data for hr. The outliers are `r outliers(df$diasbp)` for the Diastolic Blood Pressure.


### bmi 
The BMI of the individual admitted to the hospital

```{r}
summary(df$bmi)
ggplot(df, aes(x = bmi)) + geom_histogram() + geom_vline(aes(xintercept = mean(df$bmi)), color = "red")
```

There were `r length(outliers(df$bmi))` outliers found in the data for hr. The outliers are `r outliers(df$bmi)` for the BMI.


### cvd

The cvd variable refers to History of Cardiovascular Disease of the individual admitted to the hospital.
```{r}
print(table(df$cvd))
```



### afb

The afb variable refers to Atrial Fibrillation of the individual admitted to the hospital.
```{r}
print(table(df$afb))
```



### sho

The sho variable refers to Cardiogenic Shock of the individual admitted to the hospital.
```{r}
print(table(df$sho))
```



### chf

The chf variable refers to Congestive Heart Complications of the individual admitted to the hospital.
```{r}
print(table(df$chf))
```


### av3

The av3 variable refers to Complete Heart Block of the individual admitted to the hospital.
```{r}
print(table(df$av3))
```



### miord

The miord variable refers to MI Order of the individual admitted to the hospital.
```{r}
print(table(df$miord))
```


### mitype

The mitype variable refers to MI Type of the individual admitted to the hospital.
```{r}
print(table(df$mitype))
```


### year

The year variable refers to Cohort Year of the individual admitted to the hospital.
```{r}
print(table(df$year))
```



### dstat

The dstat variable refers to Discharge Status of the individual admitted to the hospital.
```{r}
print(table(df$dstat))
```


# Reviewing difference in collection groups

The data was collected in 3 different time periods. This section compares those time periods for to determine if they are similar or not.


### age

```{r}
ggplot(df, aes(x = age, fill = as.factor(year)))  + geom_density(alpha = 0.2) +  labs(x = "Age", y = "Percentage", fill = "Collection Year", title = "Density of Age") 
```

### gender

```{r}
ggplot(df, aes(x = gender, fill = as.factor(year))) + geom_bar(alpha = 0.4, position="dodge") + geom_bar(alpha = 0.4, position="dodge") +  labs(x = "Gender", y = "Occurrences", fill = "Year", title = "Comparison of Gender and Collection Year") 
```

### hr

```{r}
ggplot(df, aes(x = hr, fill = as.factor(year)))  + geom_density(alpha = 0.2) +  labs(x = "Heart Rate", y = "Percentage", fill = "Year", title = "Density of Heart Rate by Year") 
```

### sysbp

```{r}
ggplot(df, aes(x = sysbp, fill = as.factor(year)))  + geom_density(alpha = 0.2)+  labs(x = "Systolic Blood Pressure", y = "Percentage", fill = "Year", title = "Density of Systolic Blood Pressure by Year") 
```

### diasbp

```{r}
ggplot(df, aes(x = diasbp, fill = as.factor(year)))  + geom_density(alpha = 0.2) +  labs(x = "Diastolic Blood Pressure", y = "Percentage", fill = "Year", title = "Density of Diastolic Blood Pressure by Year") 
```

### bmi

```{r}
ggplot(df, aes(x = bmi, fill = as.factor(year)))  + geom_density(alpha = 0.2) +  labs(x = "Body Mass Index", y = "Percentage", fill = "Year", title = "Density of Body Mass Index by Year") 
```

### cvd

```{r}
ggplot(df, aes(x = cvd, fill = as.factor(year))) + geom_bar(alpha = 0.4, position="dodge") +  labs(x = "Had CVD", y = "Occurrences", fill = "Year", title = "Comparison of CVD and Collection Year") 
```

### afb

```{r}
ggplot(df, aes(x = afb, fill = as.factor(year))) + geom_bar(alpha = 0.4, position="dodge") +  labs(x = "Had AFB", y = "Occurrences", fill = "Year", title = "Comparison of AFB and Collection Year") 
```

### sho

```{r}
ggplot(df, aes(x = sho, fill = as.factor(year))) + geom_bar(alpha = 0.4, position="dodge") +  labs(x = "Had SHO", y = "Occurrences", fill = "Year", title = "Comparison of SHO and Collection Year") 
```

### chf

```{r}
ggplot(df, aes(x = chf, fill = as.factor(year))) + geom_bar(alpha = 0.4, position="dodge") +  labs(x = "Had CHF", y = "Occurrences", fill = "Year", title = "Comparison of CHF and Collection Year") 
```

### av3

```{r}
ggplot(df, aes(x = av3, fill = as.factor(year))) + geom_bar(alpha = 0.4, position="dodge") +  labs(x = "Had AV3", y = "Occurrences", fill = "Year", title = "Comparison of AV3 and Collection Year") 
```

### miord

```{r}
ggplot(df, aes(x = miord, fill = as.factor(year))) + geom_bar(alpha = 0.4, position="dodge") +   labs(x = "MI Order", y = "Occurrences", fill = "Year", title = "Comparison of MI Order and Collection Year") 
```

### mitype

```{r}
ggplot(df, aes(x = mitype, fill = as.factor(year))) + geom_bar(alpha = 0.4, position="dodge") + labs(x = "MI Type", y = "Occurrences", fill = "Year", title = "Comparison of MI Type and Collection Year") 
```

### dstat

```{r}
ggplot(df, aes(x = dstat, fill = as.factor(year))) + geom_bar(alpha = 0.4, position="dodge") +  labs(x = "Deceased Discharge", y = "Occurrences", fill = "Year", title = "Comparison of Deceased Discharge and Collection Year") 
```


# Reviewing difference in Life and Death

The data was collected in 3 different time periods. This section compares those time periods for to determine if they are similar or not.


### age

```{r}
ggplot(df, aes(x = age, fill = as.factor(dstat)))  + geom_density(alpha = 0.2) +  labs(x = "Age", y = "Percentage", fill = "Deceased", title = "Density of Age") 
```

### gender

```{r}
ggplot(df, aes(x = gender, fill = as.factor(dstat))) + geom_bar(alpha = 0.4, position="dodge") +  labs(x = "Gender", y = "Occurrences", fill = "Deceased", title = "Comparison of Gender and Discharge Status") 
```

### hr

```{r}
ggplot(df, aes(x = hr, fill = as.factor(dstat)))  + geom_density(alpha = 0.2)  +  labs(x = "Heart Rate", y = "Percentage", fill = "Discharge Status", title = "Density of Heart Rate by Discharge Status") 
```

### sysbp

```{r}
ggplot(df, aes(x = sysbp, fill = as.factor(dstat)))  + geom_density(alpha = 0.2)  +  labs(x = "Systolic Blood Pressure", y = "Percentage", fill = "Discharge Status", title = "Density of Systolic Blood Pressure by Discharge Status") 
```

### diasbp

```{r}
ggplot(df, aes(x = diasbp, fill = as.factor(dstat)))  + geom_density(alpha = 0.2)  +  labs(x = "Diastolic Blood Pressure", y = "Percentage", fill = "Discharge Status", title = "Density of Diastolic Blood Pressure by Discharge Status") 
```

### bmi

```{r}
ggplot(df, aes(x = bmi, fill = as.factor(dstat)))  + geom_density(alpha = 0.2) +  labs(x = "Body Mass Index", y = "Percentage", fill = "Discharge Status", title = "Density of Body Mass Index by Discharge Status") 
```

### cvd

```{r}
ggplot(df, aes(x = cvd, fill = as.factor(dstat))) + geom_bar(alpha = 0.4, position="dodge") +  labs(x = "Had CVD", y = "Occurrences", fill = "Deceased", title = "Comparison of CVD and Discharge Status") 
```

### afb

```{r}
ggplot(df, aes(x = afb, fill = as.factor(dstat))) + geom_bar(alpha = 0.4, position="dodge") +  labs(x = "Had AFB", y = "Occurrences", fill = "Deceased", title = "Comparison of AFB and Discharge Status") 
```

### sho

```{r}
ggplot(df, aes(x = sho, fill = as.factor(dstat))) + geom_bar(alpha = 0.4, position="dodge") +  labs(x = "Had SHO", y = "Occurrences", fill = "Deceased", title = "Comparison of SHO and Discharge Status") 
```

### chf

```{r}
ggplot(df, aes(x = chf, fill = as.factor(dstat))) + geom_bar(alpha = 0.4, position="dodge") +  labs(x = "Had CHF", y = "Occurrences", fill = "Deceased", title = "Comparison of CHF and Discharge Status") 
```

### av3

```{r}
ggplot(df, aes(x = av3, fill = as.factor(dstat))) + geom_bar(alpha = 0.4, position="dodge") +  labs(x = "Had AV3", y = "Occurrences", fill = "Deceased", title = "Comparison of AV3 and Discharge Status") 
```

### miord

```{r}
ggplot(df, aes(x = miord, fill = as.factor(dstat))) + geom_bar(alpha = 0.4, position="dodge") +  labs(x = "MI Order", y = "Occurrences", fill = "Deceased", title = "Comparison of MI Order and Discharge Status") 
```

### mitype

```{r}
ggplot(df, aes(x = mitype, fill = as.factor(dstat))) + geom_bar(alpha = 0.4, position="dodge") +  labs(x = "MI Type", y = "Occurrences", fill = "Deceased", title = "Comparison of MI Type and Discharge Status") 
```


## Bi-Variate Analysis

### Pearson test for Linearity

```{r}
# drop non numeric rows here
sub <- cor(df[, c("age", 'bmi','sysbp','diasbp', 'hr')])
round(sub, 2)
```


# Train Validate Test Split


```{r}
set.seed(101011) 
ind <- sample(3, nrow(df), replace = TRUE, prob = c(0.6, 0.25, 0.15)) 
train <- df[ind == 1, ]
train.label <- train[, ncol(train)]

validate <- df[ind == 2, ]
validate.label <- validate[, ncol(validate)]

test <- df[ind == 3, ]
test.label <- test[, ncol(test)]
print(table(ind))
```



# Logistic Regression


## Model Creation


### First attempt
```{r}
model <- glm(dstat ~ ., family = binomial, data = train)
print(model)
```

Confusion Matrix For Training Data
```{r}
table(round(predict(model, train, type="response")), train.label)
```
Confusion Matrix For Validation Data

```{r}
table(round(predict(model, validate, type="response")), validate.label)
```

Analysis of Variance
```{r}
anova(model, test = "Chisq")
```

### Second Attempt

```{r}

#verify the data
train <- df[ind == 1, ]
validate <- df[ind == 2, ]
test <- df[ind == 3, ]

cols <- c('age', 'sysbp', 'sho', 'dstat')
train <- train[, cols]
validate <- validate[, cols]
test <- test[, cols]
model <- glm(dstat ~ ., family = binomial, data = train)
print(model)
```

Evaluation on training data
```{r}
table(round(predict(model, train, type="response")), train.label)
```
Evaluation on validation data
```{r}
table(round(predict(model, validate, type="response")), validate.label)
```

Evaluation on test data
```{r}
table(round(predict(model, test, type="response")), test.label)
```

```{r}
plot(predict(model),residuals(model), col=c("blue")) 

lines(lowess(predict(model),residuals(model)), col=c("black"), lwd=2) 

abline(h=0, col="grey") 
```

### Third Attempt

```{r}
#verify the data
train <- df[ind == 1, ]
validate <- df[ind == 2, ]
test <- df[ind == 3, ]

model <- glm(dstat ~ ., family = binomial, data = train)
summary(step(model))
```


```{r}
cols <- c('age', 'sysbp', 'sho', 'av3', 'hr' ,'dstat')

train <- train[, cols]
validate <- validate[, cols]

model <- glm(dstat ~ ., family = binomial, data = train)
print(model)
```

Confusion Matrix For Training Data
```{r}
table(round(predict(model, train, type="response")), train.label)
```
Confusion Matrix For Validation Data
```{r}
table(round(predict(model, validate, type="response")), validate.label)
```

# References

Goldberg, R. (2001). Worcester Heart Attack Study. [CSV]. Retrieved from https://learn.umgc.edu/content/enforced/685926-027339-01-2225-GO1-9041/Dataset%20Descriptions/whas500.txt 

Saraswat (2022). Practical Guide to Logistic Regression Analysis in R. Retrieved from https://www.hackerearth.com/practice/machine-learning/machine-learning-algorithms/logistic-regression-analysis-r/tutorial/ 

SparkR. (n.d.). GLM: Generalized Linear Models. Retrieved from https://www.rdocumentation.org/packages/SparkR/versions/2.1.2/topics/glm 

Stats. (n.d.). Step Choose Model by AIC. Retrieved from https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/step 
